Estimating duration in the presence of misclassification

### 2024-03-30 cis-imperfect-testing.tex
Better notation in 6.3

\section{Likelihood modification for false negatives} \label{imperf-test:sec:modelling}

In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the likelihood derived in \cref{E-perf-test:sec:model}.
Using a simple model, in particular a constant test sensitivity, means the likelihood remains tractable.

@@

I will modify $p_{ik}$ to form $p_{ik}' = \prob(O'(j) = \vec{\nu}'_k \mid i(j) = i_k, \vec{\theta})$, taking into account false negatives.
I consider a mixture of two scenarios, whether the final test in $\sched'_{i(j)}$ is a false positive, with the mixture probability determined by the test sensitivity.

@@

First, the case when $E_j \leq r_k^{(e)}$.
In this case, the test at $r_k^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
% In this case, the test at $r_k^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_{i(j)}$.
The true negative occurs with probability 1, by the assumption of no false positives.
\begin{align}
&\prob(\vec{y}_k, E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= \prob(\vec{y}_k, l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\ % &\text{as no false positives}
&= \prob(\vec{y}_k \mid l_k^{(e)} \leq E_j \leq r_k^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
\label{imperf-test:eq:ll-ei-lt-ri}
\end{align}

Second, the case when $E_j > r_k^{(e)}$.
In this case, the test at $r_k^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
To avoid having to consider tests after $r_k^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right censored at $r_k^{(e)}$.
Taking the same approach as before:
\begin{align}
&\prob(\vec{y_k}, E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= \prob(\vec{y_k} \mid E_j > r_k^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} (1 - p_\text{sens}) S_{\vec{\theta}}(r_k^{(e)} - b_j + 2)
\label{imperf-test:eq:ll-ei-gt-ri}
\end{align}

These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the probability of observing $\vec{y_k}$ given that an infection occurs in individual $i$.
First, augment the data with $b_j$, and split into the cases just discussed:
\begin{align}
p_{ik}'
=& \prob(\vec{y_k} \mid p_\text{sens}, \vec{\theta}) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( \prob(\vec{y_k}, E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) + \prob(\vec{y_k}, E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \right) \\
  &\times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
\intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri} and take out the common factor:}
=\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
 & \times \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) + (1 - p_\text{sens}) S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \\ 
  & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}) \\
=\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
  & \times \sum_{b_j = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_k^{(e)} - b_j + 2) \right) \\
  & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
\label{imperf-test:eq:pia-prime}
\end{align}
Note that if $p_\text{sens} = 1$ then $p_{ia}' = p_{ia}$ (see \cref{perf-test:eq:pia}).

If $\psens$ is fixed (\ie has a point prior) and $p(b_j \mid \psens, \vec{\theta}) \propto 1$ (as assumed in \cref{E-perf-test}) then:
\begin{align}
p_{ik}'
&\propto \sum_{b = l_k^{(b)}}^{r_k^{(b)}} S_{\vec{\theta}}(l_k^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_k^{(e)} - b + 2).
\label{imperf-test:eq:pia-prime-constant}
\end{align}

@@

I now modify $p_{iu}$ to form $p_{iu}' = \prob(O'(j) = \emptyset \mid i(j) = i_k, \vec{\theta})$ to take into account false negatives.
Several mechanisms for episodes being undetected were considered when deriving $p_{iu}$ in \cref{E-perf-test:sec:prob-undetected}.
I now consider the additional mechanisms arising due to false negatives.
Specifically, episode $j$ could be undetected if the first test after $b_j$ is a false negative and then there are no subsequent positive tests.

This false negative would occur at the first test after the infection episode begins, on day $b_j + \tau_{\sched_{i(j)}}(b_j)$ (see \cref{E-perf-test:sec:prob-undetected}).
A false negative occurring requires that the episode has not yet ended but a negative still occurs.
The episode has not yet ended at the time of the test if $e_j = b_j + d_j - 1$ is after the test, that is the duration of the infection $d_j > \tau_{\sched_{i(j)}}(b_j)$.
Conditional on the episode having not yet ended, the test result is negative with probability $1 - \psens$.

For there to be no subsequent positive tests, all tests up until day $e_j$ are false negatives.
I assume there is a negligible probability of two false negatives.
Therefore, this can only occur if the episode ends before another test.
Denote the number of days between $b_j$ and the test following the false negative as $\tau^2_{\sched_{i(j)}}(b_j) \stackrel{\text{def}}{=} \tau_{\sched_{i(j)}}(\tau_{\sched_{i(j)}}(b_j) + 1)$.
Then, the episode ends before this test if $d_j < \tau^2_{\sched_{i(j)}}(b_j)$.

>>>>

\section{Posterior density modification for false negatives} \label{imperf-test:sec:modelling}

In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the posterior density derived in \cref{E-perf-test:sec:model}.
Using a simple model, in particular a constant test sensitivity, means calculating likelihood remains tractable.

@@

I will modify $p_{ik}$ to form $p_{ik}' = \prob(O'(j) = \vec{\nu}'_k \mid i(j) = i_k, \vec{\theta})$, taking into account false negatives.
I consider a mixture of two scenarios, whether the final test in $\sched'_{i(j)}$ is a false negative, with the mixture probability determined by the test sensitivity.

@@

First, the case when $E_j \leq r_k^{(e)}$.
In this case, the test at $r_k^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
% In this case, the test at $r_k^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_{i(j)}$.
The true negative occurs with probability 1, by the assumption of no false positives.
\begin{align}
&\prob(\vec{\nu}_k, E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= \prob(\vec{\nu}_k, l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\ % &\text{as no false positives}
&= \prob(\vec{\nu}_k \mid l_k^{(e)} \leq E_j \leq r_k^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
\label{imperf-test:eq:ll-ei-lt-ri}
\end{align}

Second, the case when $E_j > r_k^{(e)}$.
In this case, the test at $r_k^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
To avoid having to consider tests after $r_k^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right censored at $r_k^{(e)}$.
Taking the same approach as before:
\begin{align}
&\prob(\vec{\nu}_k, E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= \prob(\vec{\nu}_k \mid E_j > r_k^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} (1 - p_\text{sens}) S_{\vec{\theta}}(r_k^{(e)} - b_j + 2)
\label{imperf-test:eq:ll-ei-gt-ri}
\end{align}

These expressions can now be used to form the full replacement for $p_{ik}$, $p'_{ik} = \prob(O'(j ) = \vec{\nu}_k' \mid i(j) = i_k)$.
First, augment the data with $b_j$, and split into the cases just discussed:
\begin{align}
p_{ik}'
=& \prob(\vec{y_k} \mid p_\text{sens}, \vec{\theta}) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( \prob(\vec{y_k}, E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) + \prob(\vec{y_k}, E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \right) \\
  &\times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
\intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri} and take out the common factor:}
=\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
 & \times \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) + (1 - p_\text{sens}) S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \\ 
  & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}) \\
=\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
  & \times \sum_{b_j = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_k^{(e)} - b_j + 2) \right) \\
  & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
\label{imperf-test:eq:pia-prime}
\end{align}
Note that if $p_\text{sens} = 1$ then $p_{ik}' = p_{ik}$ (see \cref{perf-test:eq:pia}).

If $\psens$ is fixed (\ie has a point prior) and $\prob(B_j = b \mid \psens, \vec{\theta}) \propto 1$ (as assumed in \cref{E-perf-test}) then:
\begin{align}
p_{ik}'
&\propto \sum_{b = l_k^{(b)}}^{r_k^{(b)}} S_{\vec{\theta}}(l_k^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_k^{(e)} - b + 2).
\label{imperf-test:eq:pia-prime-constant}
\end{align}

@@

I now modify $p_{iu}$ to form $p_{iu}' = \prob(O'(j) = \emptyset \mid i(j) = i, \vec{\theta})$ to take into account false negatives.
Several mechanisms for episodes being undetected were considered when deriving $p_{iu}$ in \cref{E-perf-test:sec:prob-undetected}.
I now consider the additional mechanisms arising due to false negatives.
Specifically, episode $j$ could be undetected if the first test after $b_j$ is a false negative and then there are no subsequent positive tests.

This false negative would occur at the first test after the infection episode begins, on day $b_j + \tau_{\sched_{i(j)}}(b_j)$ (see \cref{E-perf-test:sec:prob-undetected}).
A false negative occurring requires that the episode has not yet ended but a negative still occurs.
The episode has not yet ended at the time of the test if $e_j = b_j + d_j - 1 \geq b_j + \tau_{\sched_{i(j)}}(b_j) + 1$, that is the duration of the infection $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
Conditional on the episode having not yet ended, the test result is negative with probability $1 - \psens$.

For there to be no subsequent positive tests, all tests up until day $e_j$ are false negatives.
I assume there is a negligible probability of two false negatives.
Therefore, this can only occur if the episode ends before another test.
Denote the number of days between $b_j$ and the test following the false negative as $\tau^2_{\sched_{i(j)}}(b_j) \stackrel{\text{def}}{=} \tau_{\sched_{i(j)}}(\tau_{\sched_{i(j)}}(b_j) + 1)$.
Then, the episode ends before this test if $d_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1$.
